% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  abstract=true]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,left=30mm,right=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% TODO: Add custom LaTeX header directives here
\usepackage{float, booktabs, siunitx, orcidlink}
\setkomafont{disposition}{\bfseries}
\providecommand{\keywords}[1]
{
  \small	
  \textbf
{\textit{Keywords---}} #1
}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={European or American? A re-examination of the transatlantic influence over the IMF},
  pdfkeywords={IMF, US hegemony, Europe},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{European or American? A re-examination of the transatlantic
influence over the IMF}
\author{Dianyi Yang~\orcidlink{0009-0004-4652-3429}\textsuperscript{}}
\date{2025-04-22}

\begin{document}
\maketitle
\begin{abstract}
This is an abstract.
\end{abstract}
\begin{center}
\keywords{IMF, US hegemony, Europe}
\end{center}


\newpage

\section{Introduction}\label{introduction}

The International Monetary Fund (IMF) and the World Bank\footnote{Familar
  name for the International Bank for Reconstruction and Development
  (IBRD), the lending arm of the later created World Bank Group.} were
established toward the end of World War II as part of the Bretton Woods
agreement in 1944 (\citeproc{ref-jensen2004}{Jensen, 2004}). While the
World Bank specialised in long-term project financing, the IMF was
initially designed for addressing the short-term balance-of-payment
crises which would otherwise undermine the fixed exchange rate
arrangements under the Bretton Woods system
(\citeproc{ref-jensen2004}{Jensen, 2004}).

One renowned fact about the IMF and its sister institution, the World
Bank, is the long-standing ``gentlemen's agreement'' of the IMF managing
director being a European, and the World Bank president being an
American (\citeproc{ref-Keating2024}{Keating, 2024};
\citeproc{ref-Weiss2023}{Weiss, 2023}), despite no such provisions in
either institution's articles of agreements
(\citeproc{ref-Fund2020}{IMF, 2020}; \citeproc{ref-WorldBank2021}{World
Bank, 2021}).

Such arrangements inspire a straightforward question that is yet
difficult to answer: who has more influence over the IMF, the US or
Europe? Whereas the US dominance in the World Bank is undisputed
(\citeproc{ref-clark2020}{Clark \& Dolan, 2020};
\citeproc{ref-wade2002}{Wade, 2002}; \citeproc{ref-woods2003}{Woods,
2003}), the case of the IMF is much more complicated. As the
\hyperref[sec-litreview]{Literature Review} section demonstrates, there
has been abundant evidence of both European and American influence on
the IMF, but short of a qualitative or quantitative comparison between
the two.

Qualitative attempts at this question may have been thwarted by the
closed-door and consensus-based style of governance of the IMF, as
epitomised by the ``gentlemen's agreement''. On the one hand, the
frequent use of conventions and consensus renders any analyses on the
formal structure of the IMF (e.g.~veto power) insufficient on its own.
On the other hand, the closed-door negotiations imply that we often
observe the outcome, but not the process of the bargaining. While
specific case studies (e.g.~Mexico and European debt crisis, see Lipscy
\& Lee (\citeproc{ref-lipscy2018}{2018}) and Hodson
(\citeproc{ref-hodson2014}{2014})) may have shed light on the influence
of the US or Europe, they rarely allow for a transatlantic comparison
and their representativeness can be questioned.

This dissertation therefore attempts to quantitatively compare the
influence of the US and Europe on the IMF.

The structure of this dissertation is as follows:
Section~\ref{sec-litreview} provides a review over the literature on the
influence of the US and Europe over the IMF.

\section{Literature Review}\label{sec-litreview}

\begin{itemize}
\tightlist
\item
  IMF chief European (\citeproc{ref-Kahler2001}{Kahler, 2001})
\end{itemize}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-clark2020}
Clark, R., \& Dolan, L. R. (2020). Pleasing the Principal: U.S.
Influence in World Bank Policymaking. \emph{American Journal of
Political Science}, \emph{65}(1), 36--51.
\url{https://doi.org/10.1111/ajps.12531}

\bibitem[\citeproctext]{ref-hodson2014}
Hodson, D. (2014). The IMF as a de facto institution of the EU: A
multiple supervisor approach. \emph{Review of International Political
Economy}, \emph{22}(3), 570--598.
\url{https://doi.org/10.1080/09692290.2014.956136}

\bibitem[\citeproctext]{ref-Fund2020}
IMF. (2020). \emph{{Articles} of {Agreement}}. International Monetary
Fund.

\bibitem[\citeproctext]{ref-jensen2004}
Jensen, N. M. (2004). Crisis, Conditions, and Capital. \emph{Journal of
Conflict Resolution}, \emph{48}(2), 194--210.
\url{https://doi.org/10.1177/0022002703262860}

\bibitem[\citeproctext]{ref-Kahler2001}
Kahler, M. (2001). \emph{Leadership selection in the major
multilaterals}. Institute for International Economics.

\bibitem[\citeproctext]{ref-Keating2024}
Keating, J. E. (2024). Why {Is} the {IMF} {Chief} {Always} a {European}?
In \emph{Foreign Policy}.
\url{https://foreignpolicy.com/2011/05/18/why-is-the-imf-chief-always-a-european/}

\bibitem[\citeproctext]{ref-lipscy2018}
Lipscy, P. Y., \& Lee, H. N.-K. (2018). The IMF As a Biased Global
Insurance Mechanism: Asymmetrical Moral Hazard, Reserve Accumulation,
and Financial Crises. \emph{International Organization}, \emph{73}(1),
35--64. \url{https://doi.org/10.1017/s0020818318000371}

\bibitem[\citeproctext]{ref-wade2002}
Wade, R. H. (2002). US hegemony and the World Bank: the fight over
people and ideas. \emph{Review of International Political Economy},
\emph{9}(2), 215--243. \url{https://doi.org/10.1080/09692290110126092}

\bibitem[\citeproctext]{ref-Weiss2023}
Weiss, M. A. (2023). Selecting the world bank president. In
\emph{Congressional Research Service}.
\url{https://sgp.fas.org/crs/row/R42463.pdf}

\bibitem[\citeproctext]{ref-woods2003}
Woods, N. (2003). \emph{The united states and the international
financial institutions: Power and influence within the world bank and
the IMF} (pp. 92--114). Oxford University PressOxford.
\url{https://doi.org/10.1093/0199261431.003.0005}

\bibitem[\citeproctext]{ref-WorldBank2021}
World Bank. (2021). \emph{Articles of agreement}.
\url{https://www.worldbank.org/en/about/articles-of-agreement}

\end{CSLReferences}




\end{document}
