---
title: European or American? A re-examination of the transatlantic influence over the IMF
date: last-modified
authors:
  - name: Dianyi Yang
    email: d.yang17@lse.ac.uk
    orcid: 0009-0004-4652-3429
    correspondence: true
format:
  #native: default
  #docx: default
  orcid-pdf:
    keep-tex: true
    number-sections: true
    include-in-header:
      - text: |
          \newcommand{\centerstart}{\begin{center}}
          \newcommand{\centerend}{\end{center}}
          \usepackage{amsthm, thm-restate}
          \newtheorem{hyp}{Hypothesis}
          \newtheorem{subhyp}{Hypothesis}[hyp]
          \renewcommand{\thesubhyp}{\thehyp\alph{subhyp}}
abstract: This is an abstract.
keywords: IMF, US hegemony, Europe
execute:
  echo: false
  warning: false
  error: false
  message: false
---

\newpage

# Introduction

The International Monetary Fund (IMF) and the World Bank[^1] were established toward the end of World War II as part of the Bretton Woods agreement in 1944 [@jensen2004]. While the World Bank specialised in long-term project financing, the IMF was initially designed for addressing the short-term balance-of-payment crises which would otherwise undermine the fixed exchange rate arrangements under the Bretton Woods system [@jensen2004].

[^1]: Familiar name for the International Bank for Reconstruction and Development (IBRD), the lending arm of the later created World Bank Group.

One renowned fact about the IMF and its sister institution, the World Bank, is the long-standing "gentlemen's agreement" of the IMF managing director being a European, and the World Bank president being an American [@keating2024; @weiss2023], despite no such provisions in either institution's articles of agreements [@fund2020; @WorldBank2021].

Such arrangements inspire a straightforward question that is yet difficult to answer: who has more influence over the IMF, the US or Europe? Whereas the US dominance in the World Bank is undisputed [@clark2020; @wade2002; @woods2003], the case of the IMF is much more complicated. As the [Literature Review](#sec-litreview) section demonstrates, there has been abundant evidence of both European and American influence on the IMF, but short of a qualitative or quantitative comparison between the two.

Qualitative attempts at this question may have been thwarted by the closed-door and consensus-based style of governance of the IMF, as epitomised by the "gentlemen's agreement". On the one hand, the frequent use of conventions and consensus renders any analyses on the formal structure of the IMF (e.g. veto power) insufficient on its own. On the other hand, the closed-door negotiations imply that we often observe the outcome, but not the process of the bargaining. While specific case studies (e.g. Mexico and European debt crisis, see @lipscy2018 and @hodson2014) may have shed light on the influence of the US or Europe, they rarely allow for a transatlantic comparison and their representativeness can be questioned.

This paper therefore attempts to quantitatively compare the influence of the US and Europe on the IMF.

The structure of this paper is as follows: @sec-litreview provides a review over the literature on the influence of the US and Europe over the IMF.

# Literature Review {#sec-litreview}

-   IMF chief European [@kahler2001]

I wrote some lit review.

I wrote sth else.

# Empirical Analysis

## Hypotheses

Since the [literature review](#sec-litreview) has outlined arguments for both US and European influence over IMF decision-making, this section turns to a quantitative comparison of their respective impacts. Accordingly, we formulate our main hypothesis:

\begin{restatable}[$H_{1}$]{hyp}{hypmain}
\label{hyp-main}
The United States and Europe exert different levels of influence on IMF lending decisions.
\end{restatable}

Moreover, prior literature has either combined the United States and Europe under the umbrella of “Western” influence on IMF lending decisions [@lipscy2018], or has examined the influence of one without accounting for the other [@oatley2004; @thacker1999]. To address this gap, we reassess the influence of the United States and Europe separately, while controlling for the presence of the other. Accordingly, we formulate the following two hypotheses:

```{=latex}
\setcounter{hyp}{2}
\begin{restatable}[$H_{2a}$]{subhyp}{hypus}
\label{hyp-us}
The United States exerts a statistically significant influence on IMF lending decisions, controlling for European influence.
\end{restatable}
\vspace{-1em} % remove vertical space
\begin{restatable}[$H_{2b}$]{subhyp}{hypeu}
\label{hyp-eu}
Europe exerts a statistically significant influence on IMF lending decisions, controlling for United States influence.
\end{restatable}
```

## Data

```{r}
dta <- haven::read_dta('data/panel_data.dta') 
```

This paper uses the dataset compiled by @lipscy2018, which extends the @barro2005 dataset through 2010. It provides panel data for 130 countries in five-year intervals from 1980 to 2010[^2], including information on IMF borrowing and participation outcomes, economic indicators, and measures of each country's relationship with the United States and major European powers.

[^2]: e.g. 1980-1985, 1985-1990.

## Dependent Variables

In line with @lipscy2018 and @barro2005, we use the following four measures of IMF lending outcomes as our dependent variables:

-   IMF loan to GDP ratio
-   IMF participation rate
-   IMF loan to GDP approval
-   Number of IMF conditions

These variables follow the definitions used in @lipscy2018. The *loan-to-GDP ratio* captures the average size of IMF lending relative to a country's GDP over each five-year interval. The *participation rate* measures the share of months within each interval during which a country was subject to an IMF programme. The *approval indicator* is a binary variable coded 1 if a new IMF programme was initiated during the interval, and 0 otherwise [see also @barro2005]. Finally, the *number of conditions* reflects the total number of performance criteria, benchmarks, and prior actions required for disbursement under IMF programmes [see also @copelovitch2010; @dreher2015].

## Results

```{r results}
#| label: tbl-results
#| tbl-cap: Comparative Effects of US and EU Influence on IMF Lending Outcomes (1980–2010)
#| tbl-pos: H
load('save/regTable.RData')
resultsTable |> tinytable::theme_tt("resize", width = 1)
```

# Case Study

# Conclusion

# References {.unnumbered}

::: {#refs}
:::