---
title: European or American? A re-examination of the transatlantic influence over the IMF
header-includes: |
  \AtBeginDocument{\title[European or American (BISA 2026)]{European or American? A re-examination of the transatlantic influence over the IMF}}
date: last-modified
date-format: |
    | DD MMMM YYYY
author:
- name: Dianyi Yang
- name: Tong Hu
- name: Zhenyi Chen
format:
  beamer:
    theme: Madrid
    link-citations: true
    include-in-header: 
      - text: |
          \definecolor{myblue}{RGB}{0, 113, 187}
          \colorlet{beamer@blendedblue}{myblue}
---

## Background (1/2)

- Bretton Woods agreement in 1944: established the International Monetary Fund (IMF) and the World Bank [@jensen2004].

  - World Bank: focuses on economic assistance for *development* projects.
  - IMF: initially focused on *exchange rate stability* and *balance-of-payment* crises.

- "Gentlemen's agreement" [@keating2024; @weiss2023]:

  - the IMF managing director being a \alert{European};
  - the World Bank president being an American.

## Background (2/2)

- IMF gradually expanded its role into *structural adjustment* and *domestic economic governance*, attaching \alert{conditions} to its loans [@barnett2004].

  - Neoclassical "sound economic" policies: fiscal austerity, trade liberalization, privatization, deregulation [@nelson2014].
  - Yet many have documented *negative* consequences [@garuda2000; @lang2020; @oberdabernig2013; @dreher2006; @przeworski2000; @casper2017; @hartzell2010].

- Neo-colonialism critique: serving the interests of the \alert{West} [@stone2011; @barnett2004; @cain2022; @chen2021; @hickel2020; @moyo2024].

## Core Research Question

But the “West” is not a monolith.

- Western actors may have divergent interests and disagree among themselves on lending decisions [@stone2011, p.51].

\begin{block}{Central Puzzle:}

Who has more influence over the IMF — the U.S. or Europe?

\end{block}

- What do you think?

## Competing Claims

- The case for U.S. dominance:

  - Hegemon of the Bretton Woods system
  - Headquarters in Washington, D.C. [@stone2011, p.57]
  - Veto power and the largest single-country vote share

- The case for European dominance:

  - European managing director
  - Larger combined voting share
  - Coordination mechanisms: SCIMF and EURIMF [@lutz2014]

- Difficulties in comparison:

  - Consensus-based governance
  - Closed-door negotiations

## Empirical Strategy (1/3)

::: callout-note
## Key assumption

Western powers are more likely to support favourable IMF lending outcomes for countries with which they maintain closer political and economic relationships.
:::

$\Rightarrow$ We proxy U.S. and European influence using measures of political and economic alignment and assess whether these are *systematically associated* with IMF lending outcomes [@barro2005; @lipscy2018].

::: callout-warning
The available data do not allow us to distinguish between active influence (e.g., lobbying or direct political pressure) and passive influence (e.g., IMF efforts to maintain good relations with powerful member states).
:::

## Empirical Strategy (2/3)

```{r}
#| label: read-data
library(tidyverse)
dta <- read_rds('data/panel_data_pca.rds')
```

- Data source: @lipscy2018

  - `r dta |> pull(country) |> unique() |> length()` countries, `r dta |> pull(panel) |> min() + 5` - `r dta |> pull(panel) |> max() + 5`

  - Focus: IMF lending as global financial insurance

  - Treats US and European ties as a single measure of "Western influence"

- Proxies for political and economic alignment

  - UN General Assembly voting similarity

  - Bilateral trade intensity

  - Banking sector exposure

## Dimension Reduction: PCA

The three alignment proxies are likely to be highly correlated. Including them separately may therefore produce **unstable coefficient estimates** and reduce statistical power.

- Solution: \alert{Principal Component Analysis (PCA)} [@dunteman2008].

PCA extracts the common variation across:

- UN voting similarity
- Trade intensity
- Banking exposure

$\Rightarrow$ Constructing a single latent measure of **U.S. alignment** and a single latent measure of **European alignment**[^1].

[^1]: We operationalize "Europe" as the United Kingdom, France, and Germany — the three largest European economies with the most institutional weight in the IMF.

## Empirical Strategy (3/3)

- Outcome Variables[^2]:

  - IMF loan to GDP ratio, IMF participation rate

  - IMF loan approval, IMF conditions

- Control Variables:

  - Proxies for borrower influence

    - Number of nationals employed as IMF economists

    - IMF quota.

  - GDP, Growth

  - Reserves, OECD

[^2]: We only include Stand-By Arrangements (SBA) and Extended Fund Facility (EFF) programmes for comparability with prior studies [@lipscy2018; @barro2005; @dreher2006].

## Model Specification

$$Y^*_{it} = \alpha + \beta_1 \text{USA}_{it} + \beta_2 \text{EUP}_{it} + \gamma^\top \mathbf{P}_{it} + \delta^\top \mathbf{Z}_{it} + \lambda_t + \epsilon_{it}$$

where:

- $\text{USA}_{it}$, $\text{EUP}_{it}$​: PCA-based alignment measures
- $\mathbf{P}_{it}$: GDP and GDP per capita controls (including quadratic terms) [@barro2005; @lipscy2018].
- $\mathbf{Z}_{it}$: macroeconomic and institutional controls
- $\lambda_t$: five-year period fixed effects.

**Estimation**

IMF lending outcomes are heavily left-censored at zero.

- \alert{Tobit} (continuous outcomes): loan size, participation rate, conditionality
- \alert{Probit} (binary outcome): loan approval

## Main Results

```{r}
#| label: fig-coef-main
coef_plot <- read_rds("save/coef_plot.rds")
coef_plot
```

## Results Interpretation

**Europe: Institutional influence**

Stronger European alignment is associated with:

- higher loan approval rates
- larger loan sizes
- greater programme participation

$\Rightarrow$ European influence appears concentrated in \alert{access} to IMF resources.

**United States: Programmatic influence**

Stronger US alignment is associated with:

- fewer IMF conditions (*but not necessarily less stringent conditions*)

$\Rightarrow$ American influence appears concentrated in the \alert{design} of IMF programmes rather than lending access.

## Possible Interpretation

**Europeans as key defenders of the "Washington Consensus"?**

- Closer ties with Europe are not associated with fewer conditions.
- One possible explanation is that European actors view fiscal discipline and structural reform as *desirable* policy outcomes rather than concessions to be relaxed for political partners.
  - Unwilling rather than unable to reduce conditionality.
  - Echoes @lutz2014, who argue that European actors have often adhered more strongly to orthodox economic principles than their U.S. counterparts.

$\Rightarrow$ European influence may facilitate access to IMF resources without softening programme discipline.

**Caveat**

\alert{This interpretation is suggestive rather than directly tested.}

What qualitative evidence or case studies might help evaluate this argument?

## Explaining Europe's influence: Colonial ties?

@stone2004 argues that Britain and France sometimes intervened on behalf of former colonies in Africa.

**Test**

Interact European alignment with African recipient status: $\text{EUP} \times \text{Africa}$

**Findings**

- European alignment remains positively associated with loan approval, loan size, and participation.
- The $\text{EUP} \times \text{Africa}$ interaction term is generally negative and statistically insignificant.

**Interpretation**

$\Rightarrow$ European influence does not appear to be driven primarily by post-colonial favouritism.

$\Rightarrow$ Consistent with a broader institutional form of influence.

## Explaining European Influence: Continental preference?

European governments may be more willing to support IMF lending to fellow European countries.

- Especially plausible given later Troika arrangements during the Eurozone crisis.

**Test**: Interaction term $\text{EUP} \times \text{European Recipient}$

**Findings**

- European effects remain positive and significant.
- $\text{EUP} \times \text{European Recipient}$ is generally does not favour fellow European borrowers.

**Interpretation**

$\Rightarrow$ European influence does not appear primarily driven by preferential treatment of European borrowers.

$\Rightarrow$ Consistent with a broader institutional form of influence.

## Thank You

\vspace{1em}

Thank you for listening!

\vspace{1.5em}

I particularly welcome feedback on:

\vspace{0.5em}

- Europe and neoliberal orthodoxy
- Case studies that could help evaluate the argument
- Alternative mechanisms or interpretations
- Relevant qualitative or historical evidence

\vspace{1em}

\begin{center}
\texttt{\small dianyi.yang@politics.ox.ac.uk}
\end{center}

## References {.unnumbered .allowframebreaks}

::: {#refs}
:::